---
title: "Introduction to bulknull"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to bulknull}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The **bulknull** package provides tools for assessing whether a non-significant bulk-level result reflects genuine absence of effect or dilution of a signal in a small cell-type subcluster. It implements two key statistics:

1. **Dilution Discordance Score (DDS)**: Posterior probability that bulk nullness reflects dilution
2. **Diagnosability Index (DI)**: Pre-data power assessment for detecting dilution

## Installation

```r
# Install from GitHub (if available)
# devtools::install_github("arora-lab/bulknull")
```

## Quick Start

Load the package and example data:

```{r setup}
library(bulknull)
data(dpp9_m8)
```

The example dataset contains measurements for DPP9 expression in the M8 (inflammatory macrophages) subcluster:

```{r}
str(dpp9_m8)
```

## Computing the Dilution Discordance Score

The DDS quantifies whether a bulk-null result could reflect dilution of a single-cell signal:

```{r}
result <- dilution_score(
  bulk_beta = dpp9_m8$bulk_beta,
  bulk_se = dpp9_m8$bulk_se,
  bulk_fdr = dpp9_m8$bulk_fdr,
  sc_zscore = dpp9_m8$sc_zscore,
  cluster_fraction = dpp9_m8$cluster_fraction,
  condition_fraction = dpp9_m8$condition_fraction,
  n_cluster = dpp9_m8$n_cluster
)
```

Print the results:

```{r}
cat("DDS Score:", result$dds_score, "\n")
cat("Interpretation:", result$summary, "\n")
cat("Applicable:", result$applicable, "\n")
cat("Note:", result$applicability_note, "\n")
```

## Interpreting the Dilution Discordance Score

The DDS is a posterior probability in [0, 1]:

- **DDS > 0.7**: Strong evidence for dilution (signal likely present in cluster, diluted out in bulk)
- **0.5 < DDS ≤ 0.7**: Weak-to-moderate evidence for dilution
- **0.3 < DDS ≤ 0.5**: Weak evidence for dilution
- **DDS ≤ 0.3**: Minimal evidence for dilution; consistent with genuine bulk null

In the DPP9/M8 case, DDS ≈ 0.48, indicating ambiguity: the result could reflect either true absence or dilution.

## Applicability Gate

The framework includes an applicability gate that requires the bulk result to be non-significant:

```{r}
# This case is applicable (bulk FDR > 0.05)
if (!result$applicable) {
  warning("This case is not applicable for DDS analysis.")
  warning(result$applicability_note)
}
```

For bulk-significant results, use standard cell-type-specific DE methods instead.

## Computing the Diagnosability Index

The DI provides a pre-data power assessment. For the DPP9/M8 case, we compute the expected
dilution effect size (mu_dilution) and then compute DI:

```{r}
# Compute mu_dilution from the components (same formula as inside dilution_score)
n_eff_sc <- dpp9_m8$condition_fraction * (1 - dpp9_m8$condition_fraction) * dpp9_m8$n_cluster
d_sc <- dpp9_m8$sc_zscore / sqrt(n_eff_sc)
d_bulk_expected <- dpp9_m8$cluster_fraction * d_sc
mu_dilution <- d_bulk_expected / dpp9_m8$bulk_se

# Compute DI
di_result <- diagnosability_index(
  mu_dilution = mu_dilution,
  alpha = 0.05,
  sided = "one"
)
```

Print the DI result:

```{r}
cat("Diagnosability Index:", di_result$di, "\n")
if (di_result$di < 0.3) {
  cat("Low power: Study design may be inadequate for dilution detection.\n")
} else if (di_result$di > 0.8) {
  cat("High power: Study design provides good power for dilution detection.\n")
} else {
  cat("Moderate power: Study design has moderate power for dilution detection.\n")
}
```

## Workflow: DDS with DI Context

Best practice is to report DDS alongside DI:

1. **Compute DI first** (pre-data or when planning study)
2. **If DI < 0.3**: Study is underpowered; report DDS but note low power
3. **If 0.3 ≤ DI ≤ 0.8**: Moderate power; interpret DDS carefully
4. **If DI > 0.8**: High power; DDS interpretation is reliable

## Example: Multiple Gene-Cluster Pairs

To analyze multiple genes and clusters, apply DDS to each pair:

```{r}
# Example: Create a simple table of inputs
input_data <- data.frame(
  gene = c("DPP9", "SPP1", "CXCL10"),
  bulk_beta = c(-0.030, 0.150, 0.200),
  bulk_se = rep(0.060, 3),
  bulk_fdr = c(0.821, 0.005, 0.001),
  sc_zscore = c(2.077, 3.245, 2.890),
  cluster_fraction = c(0.0695, 0.1200, 0.0850),
  condition_fraction = rep(0.631, 3),
  n_cluster = c(1332, 2100, 1800)
)

# Apply DDS to each row (in practice, use sapply or mapply)
# results <- apply(input_data, 1, function(row) {
#   dilution_score(
#     bulk_beta = as.numeric(row["bulk_beta"]),
#     bulk_se = as.numeric(row["bulk_se"]),
#     bulk_fdr = as.numeric(row["bulk_fdr"]),
#     sc_zscore = as.numeric(row["sc_zscore"]),
#     cluster_fraction = as.numeric(row["cluster_fraction"]),
#     condition_fraction = as.numeric(row["condition_fraction"]),
#     n_cluster = as.numeric(row["n_cluster"])
#   )
# })
```

## References

- Newman et al. (2015) Robust enumeration of cell subsets from tissue expression profiles. Nature Methods.
- Borenstein et al. Introduction to Meta-Analysis (deconvolution section on linear mixtures).

## Session Info

```{r}
sessionInfo()
```
