---
title: "Interpreting Low Diagnosability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting Low Diagnosability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Problem Statement

In typical RNA-seq studies, the **Diagnosability Index (DI) is extremely low** — often below 0.1 — even for large single-cell effects. This vignette explains why dilution claims are effectively unfalsifiable at realistic cluster fractions, and what this means for study design.

## The Barrier: Realistic Cluster Fractions

The DI depends on four inputs:
- **sc_zscore**: Hypothesised within-cluster effect size (driven by biology)
- **condition_fraction**: Case proportion in cohort (driven by study design)
- **n_cluster**: Cell count in target cluster (driven by single-cell depth)
- **bulk_se**: Measurement precision (driven by cohort size)
- **cluster_fraction**: Proportion of cluster in bulk tissue (driven by cell type composition)

Of these, **cluster_fraction is the bottleneck**. Inflammatory macrophages (M8) comprise ~5–10% of myeloid cells in lung tissue; neural progenitors comprise ~1–5% of brain tissue. These are realistic values for "small" clusters.

Let's load the package and data:

```{r setup}
library(bulknull)
data(dpp9_m8)
```

## Case Study: DPP9 in M8 (Inflammatory Macrophages)

The DPP9/M8 example has:
- **sc_zscore = 2.077409** (hypothesised donor-level IPF-vs-control z-score)
- **condition_fraction = 0.631** (63% IPF cases in cohort)
- **n_cluster = 1332** (M8 cells in single-cell data)
- **cluster_fraction = 0.069** (M8 as 6.9% of myeloid cells)
- **bulk_se = 0.062774** (bulk measurement SE for DPP9)

What DI does this achieve?

```{r di_baseline}
n_eff_sc <- dpp9_m8$condition_fraction * (1 - dpp9_m8$condition_fraction) * dpp9_m8$n_cluster
d_sc <- dpp9_m8$sc_zscore / sqrt(n_eff_sc)
d_bulk_expected <- dpp9_m8$cluster_fraction * d_sc
mu_dilution <- d_bulk_expected / dpp9_m8$bulk_se

di_result <- diagnosability_index(
  mu_dilution = mu_dilution,
  alpha = 0.05,
  sided = "one"
)

cat("DI at baseline:\n")
cat("  cluster_fraction =", dpp9_m8$cluster_fraction, "\n")
cat("  bulk_se =", dpp9_m8$bulk_se, "\n")
cat("  DI =", round(di_result$di, 4), "\n")
```

**DI ≈ 0.065** — only 6.5% power. This is severely underpowered.

## Can We Improve DI by Changing Study Design?

The `required_design()` function asks: what would we need to change to achieve a target DI?

### Option 1: Increase cluster_fraction

What if the target cluster were 50% instead of 7%?

```{r option1}
design_cf <- required_design(
  target_di = 0.8,
  sc_zscore = dpp9_m8$sc_zscore,
  condition_fraction = dpp9_m8$condition_fraction,
  n_cluster = dpp9_m8$n_cluster,
  cluster_fraction = dpp9_m8$cluster_fraction,
  bulk_se = dpp9_m8$bulk_se
)

cat("To achieve DI = 0.8 by increasing cluster_fraction:\n")
cat("  Target cluster_fraction:", round(design_cf$required_cluster_fraction_raw, 4), "\n")
cat("  Requirement: IMPOSSIBLE (cluster is limited by biology)\n\n")
cat(design_cf$scenario_fixed_cf, "\n")
```

**Result: cluster_fraction = 1.32 required (impossible, max = 1.0)** — Cell-type proportions are fixed by tissue composition.

### Option 2: Improve bulk measurement precision (decrease bulk_se)

What bulk_se would be needed?

```{r option2}
design_se <- required_design(
  target_di = 0.8,
  sc_zscore = dpp9_m8$sc_zscore,
  condition_fraction = dpp9_m8$condition_fraction,
  n_cluster = dpp9_m8$n_cluster,
  cluster_fraction = dpp9_m8$cluster_fraction,
  bulk_se = dpp9_m8$bulk_se
)

cat("To achieve DI = 0.8 by decreasing bulk_se:\n")
cat("  Target bulk_se:", format(design_se$required_bulk_se, scientific = TRUE), "\n")
cat("  Fold improvement needed:",
    round(dpp9_m8$bulk_se / design_se$required_bulk_se, 1), "x\n")
cat(design_se$scenario_fixed_se, "\n")
```

**Result: bulk_se = 0.00330 required vs. 0.0628 current** — requires ~159× larger cohort, or ~100,000 samples.

## Moderate Power is Also Unattainable

What if we aim for moderate power (DI = 0.5)?

```{r option3}
design_di50 <- required_design(
  target_di = 0.5,
  sc_zscore = dpp9_m8$sc_zscore,
  condition_fraction = dpp9_m8$condition_fraction,
  n_cluster = dpp9_m8$n_cluster,
  cluster_fraction = dpp9_m8$cluster_fraction,
  bulk_se = dpp9_m8$bulk_se
)

cat("To achieve DI = 0.5 (moderate power):\n")
cat("  Target bulk_se:", format(design_di50$required_bulk_se, scientific = TRUE), "\n")
cat("  Fold improvement needed:",
    round(dpp9_m8$bulk_se / design_di50$required_bulk_se, 1), "x\n")
cat(design_di50$scenario_fixed_se, "\n")
```

**Result: still requires ~50× larger cohort (~30,000 samples).**

## Summary: Why Dilution Claims Are Unfalsifiable

1. **Cluster fraction is fixed by biology.** You cannot change the proportion of M8 cells in lung tissue; it is what it is (5–10%).

2. **Bulk measurement precision requires massive cohort expansion.** To detect dilution with even moderate power (DI = 0.5), you need dozens of thousands of additional samples — economically infeasible for most studies.

3. **At DI < 0.3, any inference is unreliable.** The bulknull framework requires high power to distinguish dilution from true absence of effect.

## When Can We Detect Dilution?

Dilution is only reliably detectable when:
- **Large cluster fraction** (e.g., ≥30% of tissue) — but then the cluster is not "small" and may show significance at bulk level anyway
- **Massive cohort** (10,000+ samples) — few studies have this scale
- **Very large effect size** (sc_zscore ≥ 3) — requires extreme within-cluster biology

## Practical Recommendation

Instead of RNA-seq alone, combine with **spatial profiling** (Xenium, Vizgen, spatial proteomics, etc.) to:
- Map cell types to tissue location with micron-scale resolution
- Measure within-cluster effects directly at the tissue level
- Avoid dilution entirely by working at single-cell resolution

Dilution is a fundamental limitation of bulk measurement, not a solvable parameter optimization problem.

## References

- Borenstein, M., et al. (2009). Introduction to Meta-Analysis.
- Newman, A. M., et al. (2015). Robust enumeration of cell subsets from tissue expression profiles. *Nature Methods*.

## Session Info

```{r}
sessionInfo()
```
